\\Large Rising temperatures alter how and where boreal mammals move

### 2025-10-16 writing/manuscript.Rmd
corrected typo; added a ref

Global warming during the last century has resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extreme-heat events [@ipcc_climate_2023], with widespread changes in mammals' phenologies, circadian rhythms, fitness, and life histories [@botero_evolutionary_2015; @wells_life_2022; @newediuk_climate_2024; @mclellan_effect_2015; @pigeon_drivers_2016]. However, behavioural responses to heat vary substantially across species depending on body size, vagility, and activity times [@mccain_body_2014; @williams_range_2018]. For example, warming temperatures have allowed deer (*Odocoileus* spp.) to shift northward, which has led to increased density of wolves (*Canis lupus*) and higher predation on caribou (*Rangifer tarandus caribou*) [@latham_invading_2011; @barber_potential_2018; @dickie_habitat_2024]. Preparing for and responding to future changes will require a detailed understanding of not only how how and where mammals move throughout the thermal landscape but also the cascade of effects that such changes will have at the population and community levels [@cunningham_opportunity_2021].

@@

In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves; Table \@ref(tab:data-table)). Western Canada is currently experiencing accelerating and widespread but spatially heterogeneous warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [plants: @post_climate_2008; @tysor_countergradient_2025; ice: @basu_phenological_2024], and more frequent and intense extreme heat events [@zhang_increased_2023]. Decision makers must understand which areas will best sustain and protect biodiversity not only for current conditions but future decades, too. Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term. Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @slatyer_ecological_2022; @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].

@@

```{r boreal-rss, fig.cap="\\textbf{Temperature affects boreal caribou's and wolves' habitat selection and, consequently, their encounter rates.} Panel (A) shows the rasters of the three habitat variables used in the hierarchical resource selection functions, while panel (B) shows the relative selection strength (RSS) for boreal caribou and wolves, as a function of temperature. RSS values were re-centered and capped to between 0.25 and 4 ($2^{\\pm 2}$) to improve readability. Finally, panel (C) shows the (scaled) product of the RSS values in panel (B) as a proxy four encounter rates between individuals of the two species."}
knitr::include_graphics('../figures/boreal-caribou-wolf-habitat.png')
```

>>>>

Global warming during the last century has resulted in milder and shorter winters, hotter and longer summers, and a greater risk of extreme-heat events [@ipcc_climate_2023], with widespread changes in mammals' phenologies, circadian rhythms, fitness, and life histories [@botero_evolutionary_2015; @le_corre_weather_2016; @wells_life_2022; @newediuk_climate_2024; @mclellan_effect_2015; @pigeon_drivers_2016]. However, behavioural responses to heat vary substantially across species depending on body size, vagility, and activity times [@mccain_body_2014; @williams_range_2018]. For example, warming temperatures have allowed deer (*Odocoileus* spp.) to shift northward, which has led to increased density of wolves (*Canis lupus*) and higher predation on caribou (*Rangifer tarandus caribou*) [@latham_invading_2011; @barber_potential_2018; @dickie_habitat_2024]. Preparing for and responding to future changes will require a detailed understanding of not only how how and where mammals move throughout the thermal landscape but also the cascade of effects that such changes will have at the population and community levels [@cunningham_opportunity_2021].

@@

In this paper, we quantify the effects of proximal air temperature on mammalian movement behaviour, namely movement rates and habitat selection of six large-bodied terrestrial mammal species in western Canada (boreal and southern mountain caribou, cougars: *Puma concolor*, elk: *Cervus canadensis*, grizzly bears: *Ursus arctos horribilis*, mountain goats: *Oreamnos americanus*, and wolves; Table \@ref(tab:data-table)). Western Canada is currently experiencing accelerating and widespread but spatially heterogeneous warming [@dierauer_climate_2021; @kienzle_has_2018; @turner_its_2009], phenological shifts [plants: @post_climate_2008; @tysor_countergradient_2025; ice: @basu_phenological_2024; caribou: @le_corre_weather_2016], and more frequent and intense extreme heat events [@zhang_increased_2023]. Decision makers must understand which areas will best sustain and protect biodiversity not only for current conditions but future decades, too. Failing to understand how climate change will affect mammalian movement behaviour will hinder our ability to respond to the current climate crisis and make decisions that are viable in the long term. Furthermore, understanding how temperature affects individual species is the first step towards understanding how temperature impacts the community dynamics [@hegel_interacting_2010; @carbeck_adaptation_2022; @cunningham_opportunity_2021] and phenology [@bunnell_global_2011; @duncan_predicting_2012; @merkle_large_2016; @le_corre_weather_2016; @slatyer_ecological_2022; @gerlich_species_2025], and how these changes will impact mammals in the coming decades [@sobie_climate_2021; @eum_potential_2023; @weststrate_how_2024].

@@

```{r boreal-rss, fig.cap="\\textbf{Temperature affects boreal caribou's and wolves' habitat selection and, consequently, their encounter rates.} Panel (A) shows the rasters of the three habitat variables used in the hierarchical resource selection functions, while panel (B) shows the relative selection strength (RSS) for boreal caribou and wolves, as a function of temperature. RSS values were re-centered and capped to between 0.25 and 4 ($2^{\\pm 2}$) to improve readability. Finally, panel (C) shows the (scaled) product of the RSS values in panel (B) as a proxy for encounter rates between individuals of the two species."}
knitr::include_graphics('../figures/boreal-caribou-wolf-habitat.png')
```
